---
title: "In-class Exercise 2C: Emerging Hot Spot Analysis: sfdep methods"
date-modified: "last-modified"
---

This in-class exercise is not complete yet

# **Overview**

Emerging Hot Spot Analysis (EHSA) is a spatio-temporal analysis method for revealing and describing how hot spot and cold spot areas evolve over time. The analysis consist of four main steps:

-   Building a space-time cube,

-   Calculating Getis-Ord local Gi\* statistic for each bin by using an FDR correction,

-   Evaluating these hot and cold spot trends by using Mann-Kendall trend test,

-   Categorising each study area location by referring to the resultant trend z-score and p-value for each location with data, and with the hot spot z-score and p-value for each bin.

# **Getting started**

## **Installing and Loading the R Packages**

As usual, `p_load()` of **pacman** package will be used to check if the necessary packages have been installed in R, if yes, load the packages on R environment.

Five R packages are need for this in-class exercise, they are: sf, sfdep, tmap, plotly and tidyverse.

```{r}
#| eval: false
pacman::p_load(sf, sfdep, tmap, plotly, tidyverse)
```

# **The Data**

For the purpose of this in-class exercise, the Hunan data sets will be used. There are two data sets in this use case, they are:

-   Hunan, a geospatial data set in ESRI shapefile format, and

-   Hunan_GDPPC, an attribute data set in csv format.

Before getting started, reveal the content of *Hunan_GDPPC.csv* by using Notepad and MS Excel.

## **Importing geospatial data**

In the code chunk below, `st_read()` of **sf** package is used to import *Hunan* shapefile into R.

```{r}
#| eval: false
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

## **Importing attribute table**

In the code chunk below, `read_csv()` of **readr** is used to import *Hunan_GDPPC.csv* into R.

```{r}
#| eval: false
GDPPC <- read_csv("data/aspatial/Hunan_GDPPC.csv")
```

# **Creating a Time Series Cube**

Refer to this [article](https://sfdep.josiahparry.com/articles/spacetime-s3.html) to learn the basic concept of spatio-temporal cube and its implementation in sfdep package.

In the code chunk below, `spacetime()` of sfdep is used to create a spacetime cube.

```{r}
#| eval: false
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

Next, `is_spacetime_cube()` of sfdep package will be used to verify if GDPPC_st is indeed an space-time cube object. if it returns TRUE, it confirms that GDPPC_st is indeed an space-time cube.

```{r}
#| eval: false
is_spacetime_cube(GDPPC_st)
```

Since it returns TRUE, this confirms that the GDPPC_st object is indeed a space-time cube.

## **Computing Gi\***

Next, we will compute the local Gi\* statistics.

### **Deriving the spatial weights**

The code chunk below will be used to identify neighbors and to derive an inverse distance weights.

```{r}
#| eval: false
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_inverse_distance(nb, geometry,
                                  scale = 1,
                                  alpha = 1),
         .before = 1) %>%
  set_nbs("nb") %>%
  set_wts("wt")
```

Notes:

-   `activate()` of dplyr package is used to activate the geometry context

-   `mutate()` of dplyr package is used to create two new columns *nb* and *wt*.

-   Then we will activate the data context again and copy over the nb and wt columns to each time-slice using `set_nbs()` and `set_wts()`

    -   row order is very important so do not rearrange the observations after using `set_nbs()` or `set_wts()`.

Note that this dataset now has neighbors and weights for each time-slice.

```{r}
#| eval: false
head(GDPPC_nb)
```

## **Computing Gi\***

We can use these new columns to manually calculate the local Gi\* for each location. We can do this by grouping by *Year* and using `local_gstar_perm()` of sfdep package. After which, we `use unnest()` to unnest *gi_star* column of the newly created *gi_starts* data.frame.

```{r}
#| eval: false
gi_stars <- GDPPC_nb %>% 
  group_by(Year) %>% 
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>% 
  tidyr::unnest(gi_star)
```

# **Mann-Kendall Test**

With these Gi\* measures we can then evaluate each location for a trend using the Mann-Kendall test. The code chunk below uses Changsha county.

```{r}
#| eval: false
cbg <- gi_stars %>% 
  ungroup() %>% 
  filter(County == "Changsha") |> 
  select(County, Year, gi_star)
```

Next, we plot the result by using ggplot2 functions.

```{r}
#| eval: false
ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()
```

We can also create an interactive plot by using `ggploty()` of **plotly** package.

```{r}
#| eval: false
p <- ggplot(data = cbg, 
       aes(x = Year, 
           y = gi_star)) +
  geom_line() +
  theme_light()

ggplotly(p)
```

```{r}
#| eval: false
cbg %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>% 
  tidyr::unnest_wider(mk)
```

In the above result, sl is the p-value. This result tells us that there is a slight upward but insignificant trend.

We can replicate this for each location by using `group_by()` of dplyr package.

```{r}
#| eval: false
ehsa <- gi_stars %>%
  group_by(County) %>%
  summarise(mk = list(
    unclass(
      Kendall::MannKendall(gi_star)))) %>%
  tidyr::unnest_wider(mk)
```

## **Arrange to show significant emerging hot/cold spots**

```{r}
#| eval: false
emerging <- ehsa %>% 
  arrange(sl, abs(tau)) %>% 
  slice(1:5)
```

## **Performing Emerging Hotspot Analysis**

Lastly, we will perform EHSA analysis by using [`emerging_hotspot_analysis()`](https://sfdep.josiahparry.com/reference/emerging_hotspot_analysis.html) of sfdep package. It takes a spacetime object x (i.e. GDPPC_st), and the quoted name of the variable of interest (i.e. GDPPC) for .var argument. The k argument is used to specify the number of time lags which is set to 1 by default. Lastly, nsim map numbers of simulation to be performed.

```{r}
#| eval: false
ehsa <- emerging_hotspot_analysis(
  x = GDPPC_st, 
  .var = "GDPPC", 
  k = 1, 
  nsim = 99
)
```

### **Visualising the distribution of EHSA classes**

In the code chunk below, ggplot2 functions ised used to reveal the distribution of EHSA classes as a bar chart.

```{r}
#| eval: false
ggplot(data = ehsa,
       aes(x = classification)) +
  geom_bar()
```

Figure above shows that sporadic cold spots class has the high numbers of county.

### **Visualising EHSA**

In this section, you will learn how to visualise the geographic distribution EHSA classes. However, before we can do so, we need to join both *hunan* and *ehsa* together by using the code chunk below.

```{r}
#| eval: false
hunan_ehsa <- hunan %>%
  left_join(ehsa,
            by = join_by(County == location))
```

Next, tmap functions will be used to plot a categorical choropleth map by using the code chunk below.

```{r}
#| eval: false
ehsa_sig <- hunan_ehsa  %>%
  filter(p_value < 0.05)
tmap_mode("plot")
tm_shape(hunan_ehsa) +
  tm_polygons() +
  tm_borders(alpha = 0.5) +
tm_shape(ehsa_sig) +
  tm_fill("classification") + 
  tm_borders(alpha = 0.4)
```
