---
title: "In-class Exercise 2C: EHSA"
date-modified: "last-modified"
---

**NOT complete yet**

```{r}
pacman::p_load(sf, sfdep, tmap, plotly, tidyverse)
```

Time-Space Cube

space_time() of sfdep is used to create a spacetime cube.

```{r}
#| eval: false
GDPPC_st <- spacetime(GDPPC, hunan,
                      .loc_col = "County",
                      .time_col = "Year")
```

Next, `is_spacetime_cube()` of sfdep package will be used to cerify if GDPPC_st is indeed an space time cube object. if it returns TRUE , confirms that GDPPC_st is indeed an time space cube.

```{r}
#| eval: false
is_spacetime_cube(GDPPC_st)
```

```{r}
#| eval: false
GDPPC_nb <- GDPPC_st %>%
  activate("geometry") %>%
    mutate()
```

Computing Gi\*

We can use these new columns to manually calculate the local Gi\* for each location, we can do this by grouping by Year and using `local_gstar_perm()` of sfdep package. After which, we use `unnest()` to unnest `gf_star` column of the newly created `gi_star`s data frame.

```{r}
#| eval: false
gi_stars <- GDPPC_nb %>%
  group_by(Year) %>%
  mutate(gi_star = local_gstar_perm(
    GDPPC, nb, wt)) %>%
  tidyr::unnest(gi_star)
```

Performing Emerging Hotspot Analysis

need this for take-home exercise 1
